JD notes on effects

### 2022-06-17 manuscript/body.tex
Autocommit

In multivariate models, we may want to visualize the effect of a particular focal input variable. In this case, effects styled plots are preferable over the conventional prediction plots which generally focus on the \emph{total} effects. To illustrate this, we simulate a multivariate model described in \nameref{S1_Appendix}. We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables.

We generated two sets of predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial polynomial focal predictor); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable). Then compared the predictions and effects as shown in Fig~\ref{fig:pred_cubic_plots}.

\begin{figure}
\begin{center}
\includegraphics{cubic_predictors_preds.ggp.pdf}
\end{center}
\caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial. B: The focal predictor is a linear function, with a cubic polynomial non-focal predictor. In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but intersects at the mean of the focal predictor (model center) only in the case of linear focal predictor -- B. In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves will not necessarily cross at a point -- A. The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the effects for A and B, respectively; while the outer blue curves are the \BC{predictions (place holder for now)} in both cases. The central curve and trend line, in A and B, respectively; are the predicted household size based on age and wealth index, respectively. \bmb{Use different line types ...}\BC{Confirm with BMB}}
\label{fig:pred_cubic_plots}
\end{figure}

In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and all intersect at the mean-anchored confidence bands (effects), at the model center (Fig~\ref{fig:pred_cubic_plots}B). On the other hand, if the focal predictor is characterized by complex interactions or any other form of transformation, the model center is not necessarily a point in the predictor space. In this case, the mean-anchored confidence bands will be narrower than the non-anchored ones but will not necessarily cross at the model center (Fig~\ref{fig:pred_cubic_plots}A).

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, \emph{right} and \emph{left}, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.} 
\label{fig:pred_bin_plots}
\end{figure}

If there was no effect of nonlinear averaging, then we would expect the observed proportion and the average of the predicted probabilities to cross at the model center as we see in Fig~\ref{fig:pred_cubic_plots}B. However, the differences we see in Fig~\ref{fig:pred_bin_plots} are due to nonlinear averaging since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale. In particular, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the model center, and vice versa.

>>>>

To illustrate the distinction between the two approaches, we simulate a multivariate model described in \nameref{S1_Appendix}. We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables. We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal predictor); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable). In both cases the, the center point was used as the anchor for the effects. Then compared the predictions and effects as shown in Fig~\ref{fig:pred_cubic_plots}.

\begin{figure}
\begin{center}
\includegraphics{cubic_predictors_preds.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal input variable is a cubic polynomial. B: The focal input variable is linear, with polynomial non-focal input variable. The central curve and trend lines are the same for predictions and effects plots in the respective plots. However, the confidence intervals for the effects are narrower than those of the predictions in both cases. The default anchor for generating effects is the center point. For simple models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at the center point -- B. In the case of higher order interactions such as cubic polynomials, the center point is not a point but a combination of all the predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A. The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
\label{fig:pred_cubic_plots}
\end{figure}

Predictions are the standard curves (wider); they tell us what we expect to see if we know the value of the focal input variable. In multivariate models, at times, we may want to focus on the effect of the focal input variable only. Unlike prediction plot, which captures the uncertainty in the entire model, effect plot focuses on the uncertainty of the focal input variable only. In this case, effects styled plots are preferable over the conventional prediction plots which generally focus on the \emph{total} effects, and are usually narrower around the anchor.

In the absence of complex higher order interactions or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and interactions at the center point, for the effects plot. On the other hand, if the focal input variable is characterized by complex interactions or any other form of transformations, the center point is not necessarily a point in the range of focal input values. In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions.} The \emph{truth} is the prediction based on the true simulation parameters. If the focal input variable has small effect size, mean-based and whole-sample-based approaches produce different predictions -- A. In the case of strong effect size, however, the two approaches produce similar predictions -- B. In both cases, the whole-sample-based predictions are close to the truth. The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal variables and the nonlinear link function in the logistic model. Mean-based approach suffers from nonlinear averaging; and if focal input variable has small effect size, the bias is even pronounced. However, this is not the case if the non-focal input variables have small effect since the bias they induce is equally likely to be small. Since whole-sample-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned. The horizontal dashed lines correspond to the respective averages. The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin.} 
\label{fig:pred_bin_plots}
\end{figure}

If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in Fig~\ref{fig:pred_cubic_plots}B. One of possible reason for the variations we see in Fig~\ref{fig:pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale. For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
